Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-09-20 doc/invt.tex
small changes

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, wang2001pre}
and metadynamics\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, micheletti2004}
among others\cite{yan2004, kim2006, *kim2007, kim2010, junghans2014,
  langfeld2012, pellegrini2014,
  maragliano2006, *abrams2008,
  zheng2010}.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incremental local elevation of the bias potential.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
%where the updating should occur
as well as the relative updating magnitude.
%
The window function often
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in the WL algorithm,
or that of a Gaussian
in metadynamics.\cite{junghans2014}
%
We will refer to the two updating schemes
as the single-bin
and Gaussian updating schemes, respectively.
%
While the former is popular for MC simulations,\cite{wang2001,
  wang2001pre, kim2006, *kim2007}
the latter is commonly used for MD simulations
as it provides a smoother bias potential
that can be readily differentiated
to derive the bias force.\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, junghans2014}
%
In both cases,
regular updates to the bias potential
disrupts the underlying equilibrium
sampling,\cite{zhou2005, morozov2007, zhou2008}
and the convergence of the bias potential
can often benefit from a gradually decreasing
updating magnitude of the bias potential.

@@

We will then show by comparing the errors
of simulations under different updating schemes
that the single-bin updating scheme
is optimal for very long simulations.
%without a priori assumption of the smoothness of the PMF.
%
A similar but partial asymptotic optimality applies to
a class of bandpass updating schemes
that flatten only
a few long-range distribution modes.
%
Compared with the Gaussian updating scheme,
bandpass updating schemes have the advantage of allowing
simpler control of the updating magnitude
for optimal convergence.

@@

We present the methods
in the following order.
%
In Sec. \ref{sec:background},
we briefly review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we derive the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
(as in the WL algorithm)
by proving the optimality
of the known inverse-time formula.
%
We extend the method to the general case
of a multiple-bin updating scheme
that encompasses the Gaussian updating scheme used in metadynamics
in Sec. \ref{sec:multiple-bin}.
%
Then, we compare different updating schemes
and show that the single-bin updating scheme
and a class of generalized bandpass updating schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.
%
Finally,
we use the bandpass updating schemes
to build a class of adaptive FDS algorithms
with polynomial bias potentials
in Sec. \ref{sec:bandpass_poly}.

@@

Sampling a multimodal distribution with deep basins
is often challenging in regular MC and MD simulations.
%
To promote transitions among the basins,
we can apply a bias potential to alter
the intrinsic distribution to a smoother one, $\rho_i$.
%
We will broadly define
a flat-distribution-sampling (FDS) simulation
as a biased simulation
targeting a flat $\rho_i$ (i.e. $\rho_i \equiv 1/n$ for $n$ bins)
or a nearly flat one\cite{
dayal2004, *trebst2004, barducci2008, singh2011}.
%
Since the equilibrium distribution
under a bias potential, $u_i$, is given by
%
\begin{equation}
  \pi_i
  %=
  %C_i \, p_i \, e^{-u_i}
  =
  \frac{                p_i \, e^{-u_i} }
       { \sum_{j = 1}^n p_j \, e^{-u_j} }
  \propto
  p_i \, e^{-u_i}
  ,
  \notag
  %\label{eq:pi_p_V}
\end{equation}
%
we wish to find a bias potential that allows
the resulting $\pi_i$ to coincide
with the desired $\rho_i$.
%
Upon convergence, the bias potential would satisfy
%
\begin{equation}
  u_i \to \ln \frac{p_i}{\rho_i} + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
and the PMF, $-\ln p_i$, can be readily
deduced from the bias potential. % $u_i$.

@@

Below we will focus on a more recent class of adaptive FDS methods,
including the WL algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{huber1994, laio2002, *laio2008, *barducci2011, *sutto2012},
which
continuously and incrementally update the bias potential.
%
%\subsubsection{Updating schemes}
%
In the WL algorithm\cite{wang2001, wang2001pre},
the bias potential, $u_i(t)$, is updated
at each time step $t$ as
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { \rho_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as
the \emph{single-bin} (updating) scheme,
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general
\emph{multiple-bin} (updating) scheme,
%including the Gaussian updating scheme used in metadynamics,
neighboring bins are also updated:
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { \rho_{i(t)} }
  .
  \label{eq:mbin_update}
\end{equation}
%
A common example is the Gaussian updating scheme used in metadynamics,
with $w_{ij}$ being roughly a Gaussian centered on bin $j = i(t)$.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ denotes the number of updating steps.

@@

While adaptive FDS simulations
can often sample the desired distribution,
frequent updates to the bias potential
disrupt the underlying equilibrium sampling
and introduce artificial errors that need to be
reduced by gradually decreasing the updating magnitude.\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, morozov2009, zhou2008, min2007,
  poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}

@@

In this section,
we derive the optimal $\alpha(t)$
for the single-bin updating scheme,
Eq. \eqref{eq:wl_update},
as a preparation
for the more general multiple-bin scheme.
%
We will first express the error of
the bias potential, $u_i(t)$,
as a functional of $\alpha(t)$,
and then minimize it by variation.
%

@@

We can split the histogram into
a deterministic averaged part, $\langle f_i(t) \rangle$,
and a random fluctuating part, $\xi_i(t)$,
%
\begin{equation}
  f_i(t) =
  \langle f_i(t) \rangle
  +
  \xi_i(t)
  .
  \notag
  %\label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consisting of
many copies of similar simulations
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For a sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle f_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { \rho_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { \rho_i }
  =
  \frac{                          e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n \rho_j \, e^{-v_{*j}(t)} }
  %\notag
  %\\
  %&
  \approx
  1 - v_{*i}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { \rho_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n \rho_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n \rho_j \, v_{*j} }
.
$$
}%
%
Deducting the bin average
and noticing that
$\bav{f}(t) = \sum_{i=1}^n \rho_i [h_i(t)/\rho_i] = \sum_{i=1}^n \delta_{i, i(t)} = 1$,
we get
$\langle f_{*i}(t) \rangle = \langle f_i(t) \rangle - 1 \approx - v_{*i}(t)$,
%
%\begin{align*}
%  \langle f_{*i}(t) \rangle
%  \equiv
%  \langle f_i(t) \rangle -
%  \bav{f}(t)
%  =
%  \langle f_i(t) \rangle - 1
%  \approx
%  - v_{*i}(t)
%  ,
%\end{align*}
%
and
%
\begin{equation}
  f_{*i}(t) \approx - v_{*i}(t) + \xi_{*i}(t)
  .
  \label{eq:sh_ave}
\end{equation}

@@

From Eqs. \eqref{eq:error_def},
\eqref{eq:vi_solution}, and
\eqref{eq:Gi_def},
we can write the total error at the end of period $T$
as a sum of a residual error, $E_R(T)$ (from the decay of the initial error)
and an asymptotic error, $E_A(T)$:
%
\begin{align}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T)
  ,
  \label{eq:error_tot}
\end{align}
%
where
\begin{align}
  \Err_R(T)
  &= \Err(0) \, e^{-2 \, q(T)}
  ,
  \label{eq:ER_sbin}
  \\
  \Err_A(T)
  &=
  \sum_{i=1}^n \rho_i
    \int_0^T\!\!\!\int_0^T
    \!\!
    \langle
      \xi_{*i}(t) \xi_{*i}(t')
    \rangle
    \dot \theta\bigl( q(t) \bigr)
    \dot \theta\bigl( q(t') \bigr)
    \, dt dt'
  \notag \\
  &= \Gamma \, \int_0^T \dot \theta^2\bigl( q(t) \bigr) \, dt
  ,
  \label{eq:EA_sbin}
\end{align}
with
%$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.
$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i
\approx \sum_{i=1}^n \rho_i \int_{-\infty}^{\infty} \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt$.
%\begin{equation}
%  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i
%  =
%  \sum_{i=1}^n \rho_i \int_0^T
%    \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt
%  .
%  \notag%\label{eq:Gamma_from_G}
%\end{equation}

@@

We will first review the spectral decomposition
of the updating matrix, $\mathbf w$,
which helps establish the convergence criterion
and the decomposition of error into
independent eigenmodes.
%
The matrix, $\mathbf w$, is subject to several conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\pmb\rho = (\rho_1, \dots, \rho_n)$,
the growth rate of the bias potential,
${\dot u}_i(t)$
in Eq. \eqref{eq:ut_diffeq_mbin},
should be a constant of $i$
if $h_j(t)$ were the same as $\rho_j$,
i.e.
$\sum_{j=1}^n w_{ij}$ should be a constant
to allow the bias potential to grow uniformly
in the asymptotic regime.
%
If the growth rate is positive,\footnote{Although
  in general the growth rate is not necessarily positive,
  it is often the largest eigenvalue
  for ``natural'' updating matrices considered
  in this study.}
we can, by a proper overall scaling of $\mathbf w$,
write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
for cases considered here.
%
Below we will %take advantage of the analogy and
borrow techniques used
in studying transition matrices\cite{vankampen},
but limit ourselves to properties that are unaffected
by the admission of negative elements into $\mathbf w$.

@@

The above symmetry implies the existence of
a left eigenvector, $\pmb \rho$,
corresponding to the uniform right eigenvector
associated with Eq. \eqref{eq:w_sumj},
with eigenvalue $1$:
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j
  .
  \notag
  %\label{eq:w_balance}
\end{equation}
%
We will label this eigenvector by index $k = 0$,
and
%
\begin{equation}
  \phi_{0i} = \rho_i,
\label{eq:eigenmode0}
\end{equation}
%
with $\lambda_0 = 1$.
%
Clearly, Eq. \eqref{eq:eigenmode0}
satisfies the normalization condition
given by Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 0$,
while the orthogonality demands
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \delta_{k0}
  .
\label{eq:ortho0}
\end{equation}

@@

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
\begin{equation}
  {\tilde v}_k(t + 1) =
  {\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
  {\tilde f}_k(t)
  ,
  \label{eq:vkupdate}
\end{equation}
%
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde f}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \xi}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where
we have used Eq. \eqref{eq:xtstar} and
the transformed version of Eq. \eqref{eq:sh_ave}
in the second step.

@@

\item \label{step:Gamma}
Estimate $\Gamma_k$'s
(cf. Appendix \ref{sec:Gamma_measure})
and the initial bias $\epsilon_k$.\footnote{The
  value of $\tilde u_k$ can be estimated
  from the histogram-corrected bias potential
  by Eq. \eqref{eq:uhatav},
  although it often leads to overestimated
  $\epsilon_k$'s for large-$k$ modes.
}

@@

A commonly-used multiple-bin updating scheme in metadynamics
is the Gaussian updating scheme,
whose updating matrix can be characterized by
a rigid Gaussian function centered on the current bin
(with some boundary adjustment).
%to satisfy the constraints discussed in Sec. \ref{sec:updating-matrix},
%see Fig. \ref{fig:mat}(a) for an example.
%
In terms of the continuous collective variable, $z = i \, \delta z$,
the eigenvalues of this scheme are given by\cite{bussi2006}
\begin{align}
  \lambda_k
  =
  \exp\left(
        -
        \frac{ \pi^2 \, \sigma_z^2 \, k^2 }
             { 2 \, \Delta z^2 }
      \right)
  ,
  %\notag
  \label{eq:lambda_Gaussian}
\end{align}
where $\sigma_z$ is the width of the Gaussian
and $\Delta z = n \, \delta z \, g/2$
is the effective span
with $g = 1$ for a periodic variable or $2$ otherwise
(cf. Appendix \ref{sec:Gaussian_math}).

@@

By morphing the window function
but preserving the translational invariance,
we can design a homogeneous updating scheme
that is more convenient in error control
than the Gaussian updating scheme.
%
%whose updating matrix can be
%characterized by a rigid window function with a floating center,
%
As shown in Appendix \ref{sec:homo},
the homogeneous updating schemes share
the same set of orthonormal eigenvectors, $\phi_{ki}$,
which are cosine and sine functions,
and differ only by the eigenvalues.
%
Thus, the window function is completely determined
by the eigenvalues, and the updating matrix
is given by\cite{bussi2006}
%
\begin{equation}
  w_{ij}
  =
  \frac{1}{\rho_i} \sum_{k=0}^{n - 1}
  \lambda_k \, \phi_{ki} \, \phi_{kj}
  .
  \label{eq:w_from_phi}
\end{equation}
%
%We discuss this strategy in detail below.
As shown below,
the optimized updating scheme
allows the optimal schedule to coincide with
the inverse-time formula,
and thus spare the need to
go through the general procedure in Sec. \ref{sec:procedure}.

@@

\item
  Attempt a transition to another umbrella
  with the acceptance probability given
    by Eq. \eqref{eq:transprob_st}.\cite{*[{The
    intra-umbrella sampling step and
    the inter-umbrella transition step
    can be executed at different frequencies
    without affecting the correctness of the algorithm.
    Although some prefers mixing the two steps
    in a random order,
    the sequential execution adopted here
    is also valid,
    as discussed in }] [{}]
    manousiouthakis1999}

>>>>

Many efficient flat-distribution-sampling (FDS) techniques
are based on adaptive construction of the bias potential,
e.g. the Wang-Landau (WL) algorithm\cite{
  wang2001, wang2001pre}
and metadynamics\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, micheletti2004}
among others\cite{yan2004, kim2006, *kim2007, kim2010, junghans2014,
  langfeld2012, pellegrini2014,
  maragliano2006, *abrams2008,
  zheng2010}.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incremental local elevation of the bias potential.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
%where the updating should occur
as well as the relative updating magnitude.
%
The window function often
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in the WL algorithm,
or that of a Gaussian
in metadynamics.\cite{junghans2014}
%
We will refer to the two updating schemes
as the single-bin
and Gaussian updating schemes, respectively.
%
While the former is popular for MC simulations,\cite{wang2001,
  wang2001pre, kim2006, *kim2007}
the latter is commonly used for MD simulations
as it provides a smoother bias potential
that can be readily differentiated
to derive the bias force.\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, junghans2014}
%
In both cases,
regular updates to the bias potential
disrupts the underlying equilibrium
sampling,\cite{zhou2005, morozov2007, zhou2008}
and the convergence of the bias potential
can often benefit from a gradually decreasing
updating magnitude of the bias potential.

@@

We will then show by comparing the errors
of simulations under different updating schemes
that the single-bin updating scheme
is optimal for very long simulations.
%without a priori assumption of the smoothness of the PMF.
%
A similar but partial asymptotic optimality applies to
a class of bandpass updating schemes
that flatten only
a few long-range distribution modes.
%
Compared with the Gaussian updating scheme,
bandpass updating schemes allow
a simple protocol to optimally control the updating magnitude,
which is borrow from the single-bin updating scheme.

@@

We present the methods
in the following order.
%
In Sec. \ref{sec:background},
we briefly review the basics of FDS
and introduce our notations.
%
Then, in Sec. \ref{sec:single-bin},
we derive the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
(as in the WL algorithm)
by proving the optimality
of the known inverse-time formula.
%
We extend the method to the general case
of a multiple-bin updating scheme
that encompasses the Gaussian updating scheme used in metadynamics
in Sec. \ref{sec:multiple-bin}.
%
Then, we compare different updating schemes
and show that the single-bin updating scheme
and a class of generalized bandpass updating schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.
%
Finally,
we use the bandpass updating schemes
to build a class of adaptive FDS algorithms
with polynomial bias potentials
in Sec. \ref{sec:bandpass_poly}.

@@

Sampling a multimodal distribution with deep basins
is often challenging in regular MC and MD simulations.
%
To promote transitions among the basins,
we can apply a bias potential to alter
the intrinsic distribution to a smoother one, $\rho_i$.
%
We will broadly define
a flat-distribution-sampling (FDS) simulation
as a biased simulation
targeting a flat $\rho_i$ (i.e. $\rho_i \equiv 1/n$ for $n$ bins)
or a nearly flat one\cite{
dayal2004, *trebst2004, barducci2008, singh2011}.
%
Since the equilibrium distribution
under a bias potential, $u_i$, is given by
%
\begin{equation}
  \pi_i
  %=
  %C_i \, p_i \, e^{-u_i}
  =
  \frac{                p_i \, e^{-u_i} }
       { \sum_{j = 1}^n p_j \, e^{-u_j} }
  \propto
  p_i \, e^{-u_i}
  ,
  \label{eq:pi_p_V}
\end{equation}
%
we wish to find a bias potential that allows
the resulting $\pi_i$ to coincide
with the desired $\rho_i$.
%
Upon convergence, the bias potential would satisfy
%
\begin{equation}
  u_i \to \ln \frac{p_i}{\rho_i} + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
and the PMF, $-\ln p_i$, can be readily
deduced from the bias potential. % $u_i$.

@@

Below we will focus on a more recent class of adaptive FDS methods,
including the WL algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{huber1994, laio2002, *laio2008, *barducci2011, *sutto2012},
which
continuously and incrementally update the bias potential.
%
%\subsubsection{Updating schemes}
%
In the WL algorithm\cite{wang2001, wang2001pre},
the bias potential, $u_i(t)$, is updated
at each time step $t$ as
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { \rho_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as
the \emph{single-bin} updating scheme,
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general
\emph{multiple-bin} updating scheme,
%including the Gaussian updating scheme used in metadynamics,
the bias potential at neighboring bins is also updated:
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { \rho_{i(t)} }
  .
  \label{eq:mbin_update}
\end{equation}
%
A common example is the Gaussian updating scheme used in metadynamics,
with $w_{ij}$ being roughly a Gaussian centered on bin $j = i(t)$.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ denotes the number of updating steps.

@@

While adaptive FDS simulations
can often sample the desired distribution,
frequent updates to the bias potential
disrupt the underlying equilibrium sampling
and introduce artificial errors, which can be
reduced by gradually decreasing the updating magnitude.\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, morozov2009, zhou2008, min2007,
  poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}

@@

In this section,
we derive the optimal schedule, $\alpha(t)$,
for the single-bin updating scheme,
Eq. \eqref{eq:wl_update},
as a preparation
for the more general multiple-bin case.
%
We will first express the error of
the bias potential, $u_i(t)$,
as a functional of $\alpha(t)$,
and then minimize it by variation.
%

@@

We can split the $f_i(t)$ into
a deterministic averaged part, $\langle f_i(t) \rangle$,
and a random fluctuating part, $\xi_i(t)$,
%
\begin{equation}
  f_i(t) =
  \langle f_i(t) \rangle
  +
  \xi_i(t)
  .
  \notag
  %\label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consisting of
many copies of similar simulations
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For a sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
[cf. Eq. \eqref{eq:pi_p_V}]
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle f_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { \rho_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { \rho_i }
  =
  \frac{                          e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n \rho_j \, e^{-v_{*j}(t)} }
  %\notag
  %\\
  %&
  \approx
  1 - v_{*i}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion,
and used Eq. \eqref{eq:x_def} in the last step.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { \rho_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n \rho_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n \rho_j \, v_{*j} }
.
$$
But
$$
\sum_{ j = 1 }^n \rho_j \, v_{*j}
=
\sum_{ j = 1 }^n \rho_j \, v_j - \sum_{j=1}^n \bav{v}  = 0.
$$
}%
%
Deducting the bin average
and noticing that
$\bav{f}(t) = \sum_{i=1}^n \rho_i [h_i(t)/\rho_i] = \sum_{i=1}^n \delta_{i, i(t)} = 1$,
we get
$\langle f_{*i}(t) \rangle = \langle f_i(t) \rangle - 1 \approx - v_{*i}(t)$,
%
%\begin{align*}
%  \langle f_{*i}(t) \rangle
%  \equiv
%  \langle f_i(t) \rangle -
%  \bav{f}(t)
%  =
%  \langle f_i(t) \rangle - 1
%  \approx
%  - v_{*i}(t)
%  ,
%\end{align*}
%
and
%
\begin{equation}
  f_{*i}(t) \approx - v_{*i}(t) + \xi_{*i}(t)
  .
  \label{eq:sh_ave}
\end{equation}

@@

From Eqs. \eqref{eq:error_def} and
\eqref{eq:vi_solution},
we can write the total error at the end of period $T$
as a sum of a residual error, $E_R(T)$ (from the decay of the initial error)
and an asymptotic error, $E_A(T)$:
%
\begin{align}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T)
  ,
  \label{eq:error_tot}
\end{align}
%
where
\begin{align}
  \Err_R(T)
  &= \Err(0) \, e^{-2 \, q(T)}
  ,
  \label{eq:ER_sbin}
  \\
  \Err_A(T)
  &=
  \sum_{i=1}^n \rho_i
    \int_0^T\!\!\!\int_0^T
    \!\!
    \langle
      \xi_{*i}(t) \xi_{*i}(t')
    \rangle
    \dot \theta\bigl( q(t) \bigr)
    \dot \theta\bigl( q(t') \bigr)
    \, dt dt'
  \notag \\
  &= \Gamma \, \int_0^T \dot \theta^2\bigl( q(t) \bigr) \, dt
  .
  \label{eq:EA_sbin}
\end{align}
In Eq. \eqref{eq:EA_sbin},
we have used Eq. \eqref{eq:Gi_def} and introduced
%$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.
$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i
\approx \sum_{i=1}^n \rho_i \int_{-\infty}^{\infty} \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt$.
%\begin{equation}
%  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i
%  =
%  \sum_{i=1}^n \rho_i \int_0^T
%    \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt
%  .
%  \notag%\label{eq:Gamma_from_G}
%\end{equation}

@@

We will first review the spectral decomposition
of the updating matrix, $\mathbf w$,
which helps establish the convergence criterion
and the decomposition of error into
independent eigenmodes.
%
The matrix, $\mathbf w$, is subject to several conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\pmb\rho = (\rho_1, \dots, \rho_n)$,
the growth rate of the bias potential,
${\dot u}_i(t)$
in Eq. \eqref{eq:ut_diffeq_mbin},
should be a constant of $i$
if $h_j(t)$ were the same as $\rho_j$,
i.e.
$\sum_{j=1}^n w_{ij}$ should be a constant
to allow the bias potential to grow uniformly
in the asymptotic regime.
%
If the growth rate is positive,\footnote{Although
  in general the growth rate is not necessarily positive,
  it is often the largest, hence positive, eigenvalue
  for ``natural'' updating matrices considered
  in this study.}
we can, by a proper overall scaling of $\mathbf w$,
write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
for cases considered here.
%
Below we will %take advantage of the analogy and
borrow techniques used
in studying transition matrices\cite{vankampen},
but limit ourselves to properties that are unaffected
by the admission of negative elements into $\mathbf w$.

@@

The above symmetry implies the existence of
a left eigenvector, $\pmb \rho$,
corresponding to the uniform right eigenvector
associated with Eq. \eqref{eq:w_sumj},
with eigenvalue $1$:
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j
  .
  \notag
  %\label{eq:w_balance}
\end{equation}
%
We will label this eigenvector by index $k = 0$,
and
%
\begin{equation}
  \phi_{0i} = \rho_i,
\label{eq:eigenmode0}
\end{equation}
%
with $\lambda_0 = 1$.
%
Clearly, Eq. \eqref{eq:eigenmode0}
satisfies the normalization condition
given by Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 0$,
while the orthogonality between this eigenvector and others demands
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \delta_{k0}
  .
\label{eq:ortho0}
\end{equation}

@@

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme,
Eq. \eqref{eq:mbin_update} [using $u_i(t)$ instead of $v_i(t)$],
as
%
\begin{equation}
  {\tilde v}_k(t + 1) =
  {\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
  {\tilde f}_k(t)
  ,
  \label{eq:vkupdate}
\end{equation}
%
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde f}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \xi}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where
we have used Eq. \eqref{eq:xtstar} and
the transformed version of Eq. \eqref{eq:sh_ave}
in the second step.

@@

\item \label{step:Gamma}
Estimate $\Gamma_k$'s
(cf. Appendix \ref{sec:Gamma_measure})
and the initial bias $\epsilon_k$.\footnote{The
  value of $\tilde u_k$ can be estimated
  from the histogram-corrected bias potential
  by Eq. \eqref{eq:uhatav},
  although the approach often overestimates the
  $\epsilon_k$'s for large-$k$ modes.
}

@@

A commonly-used multiple-bin updating scheme in metadynamics
is the Gaussian updating scheme,
whose updating matrix can be characterized by
a rigid Gaussian function centered on the current bin
(with some boundary adjustment).
%to satisfy the constraints discussed in Sec. \ref{sec:updating-matrix},
%see Fig. \ref{fig:mat}(a) for an example.
%
In terms of the continuous collective variable, $z$,
the eigenvalues of this scheme are given by\cite{bussi2006}
\begin{align}
  \lambda_k
  =
  \exp\left(
        -
        \frac{ \pi^2 \, \sigma_z^2 \, k^2 }
             { 2 \, \Delta z^2 }
      \right)
  ,
  %\notag
  \label{eq:lambda_Gaussian}
\end{align}
where $\sigma_z$ is the width of the Gaussian
and $\Delta z = n \, \delta z \, g/2$
is the effective span
with $g = 1$ for a periodic variable or $2$ otherwise
(cf. Appendix \ref{sec:Gaussian_math}).

@@

By morphing the window function
but preserving the translational invariance,
we can design a homogeneous updating scheme
that is more convenient for error control
than the Gaussian updating scheme.
%
%whose updating matrix can be
%characterized by a rigid window function with a floating center,
%
As shown in Appendix \ref{sec:homo},
the homogeneous updating schemes share
the same set of orthonormal eigenvectors, $\phi_{ki}$,
which are cosine and sine functions,
and differ only by the eigenvalues.
%
Thus, the window function is completely determined
by the eigenvalues, and the updating matrix
is given by\cite{bussi2006}
%
\begin{equation}
  w_{ij}
  =
  \frac{1}{\rho_i} \sum_{k=0}^{n - 1}
  \lambda_k \, \phi_{ki} \, \phi_{kj}
  .
  \label{eq:w_from_phi}
\end{equation}
%
%We discuss this strategy in detail below.
As shown below,
the optimized updating scheme
allows the optimal schedule to coincide with
the inverse-time formula,
and thus spare the need to
go through the general procedure in Sec. \ref{sec:procedure}.

@@

\item
  Attempt a transition to another umbrella
  with the acceptance probability given
    by Eq. \eqref{eq:transprob_st}.\cite{*[{The
    intra-umbrella sampling step and
    the inter-umbrella transition step
    can be executed at different frequencies
    without affecting the correctness of the algorithm.
    Although a random mixing of the two steps
    is sometimes preferred,
    in a random order,
    the sequential execution adopted here
    is also valid,
    as discussed in }] [{}]
    manousiouthakis1999}
